Solution to ``Chapter 2: Basic tail and concentration bounds"

### 2020-08-30 skills/math/tail_bound/tailbound_solution.tex
tail_bound
tail bound

	To evaluate $I_1$, we solve $1 = C \exp \of{-\frac{t^2}{4 \nu^2}} $, and the minimization term becomes
	\begin{align}
		\min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} = \begin{cases}
			\quad 1 & \text{when } t < 2\nu \sqrt{\log C},\\
			\quad C \exp \of{-\frac{t^2}{4 \nu^2}} & \text{when } t \geq 2\nu \sqrt{\log C}.
		\end{cases}
	\end{align}
	Therefore, let $ y = \frac{t}{2\nu} - \sqrt{\log C}$, we obtain $I_1$ ,
	\begin{align}
		I_1 &= \int_{0}^{ 2\nu \sqrt{\log C}} 1 dt + \int_{2\nu \sqrt{\log C}}^{+\infty} C \exp \of{-\frac{t^2}{4 \nu^2}} dt\\
		&=  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2 - 2y \sqrt{\log C}} dy\\
		&\leq  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2} dy.
	\end{align}
	By Gaussian integral $\int_{0}^{+\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$, we obtain $I_1 \leq 2\nu (\sqrt{\pi} + \sqrt{\log C})$. Similarly, we evaluate $I_2$ as follows 
	\begin{align}
		I_2 &= \int_{0}^{4B \log C} 1 dt + \int_{4B \log C}^{+\infty} C \exp \of{-\frac{t}{4B}} dt\\
		 &= 4B (\log C + 1).
	\end{align}
	Hence, we obtain the expectation of $Z$,
	\begin{equation}
		\bbE[Z] = I_1 + I_2 \leq  2\nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (\log C + 1).
	\end{equation}

@@

The lower bound of $\bbE[Z_n] $ is
\begin{equation}\label{eq:e2112}
	\bbE[Z_n] \geq (1- 1/e) \sqrt{\sigma^2 \log n} , \quad \text{for all }n \geq 6.
\end{equation}
\end{lem}

@@

\begin{proof}
	%First, we prove the equation~\eqref{eq:e2111}.

	\vspace{0.2cm}
	Consider the tail probability of $Z_n$. We have
	\begin{equation}
		\bbP[Z_n \geq t] = \bbP[\max_{i \in [n]} |X_i| \geq t] = 1 - \bbP[\max_{i \in [n]} |X_i| < t] = 1 - (1- \bbP[|X_1| \geq t])^n,
	\end{equation}
	where the last equality follows from the independence of $\{X_i\}_{i=1}^n$. By Bernoulli's inequality, we have
	\begin{equation}
		(1- \bbP[|X_1| \geq t])^n \geq 1 - n \bbP[|X_1| \geq t], \quad \text{for all } t >0.
	\end{equation} 
	Therefore, the expectation of $Z_n$ follows
	\begin{equation}\label{eq:e211pro1}
		\bbE[Z_n] = \int_{0}^{+\infty} \bbP[Z_n \geq t] dt \leq  c +  \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt, \quad \text{for all } c > 0.
	\end{equation}
	 Since the tail bound $\bbP[U \geq t] \leq \sqrt{\frac{2}{\pi}} \frac{1}{t} e^{-t^2 /2}$ holds for all standard normal random variable $U$, the tail bound of $|X_1|$ satisfies
	 \begin{equation}\label{eq:e211proupper}
	 	\bbP[|X_1| \geq t] \leq 2 \sqrt{\frac{2}{\pi}} \frac{\sigma}{t} e^{-\frac{t^2}{2\sigma^2}}.
	 \end{equation}
	 Hence, let $ u = \frac{ t}{\sigma}$, the last integral in equation~\eqref{eq:e211pro1} is upper bounded as
	 \begin{align}
	 	 \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt &\leq \frac{2n \sigma}{c} \sqrt{\frac{2}{\pi}} \int_{c}^{+\infty} e^{-\frac{t^2}{2\sigma^2} }dt \\  
	 	 %& = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} e^{-\frac{u^2}{2}} dt \\
	 	 & \leq \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} u e^{-\frac{u^2}{2}} dt \\
	 	 & = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}  e^{-\frac{c^2}{2\sigma^2}}. \label{eq:e211pro2}
	 \end{align}
	 For all $n \geq 2$, let $c = \sqrt{2\sigma^2 \log n}$ and plug the $c$ into equations~\eqref{eq:e211pro1} and \eqref{eq:e211pro2}. Then, we obtain the result
	 \begin{equation}
	 	\bbE[Z_n] \leq \sqrt{2\sigma^2 \log n} + \frac{2\sigma}{\sqrt{2\log n}}   \sqrt{\frac{2}{\pi}}  \leq \sqrt{2\sigma^2 \log n} + \frac{4\sigma}{\sqrt{2\log n}}, \quad \text{for all }n \geq 2.
	 \end{equation}

@@

@@

\begin{rmk}
	The absolute Gaussian maxima $Z_n$ concentrates around $\sigma \sqrt{\log n}$ for large $n$. Standard Gaussian variables concentrate around 0, while the maxima of the $n$ standard Gaussian variables is no longer a Gaussian variables. The scale of Gaussian maxima scales as $\sqrt{n}$, however, the sum of $n$ i.i.d.\ Gaussian variables scales as $n$.
\end{rmk}

\subsection{Exercise 2.12}
\begin{lem}[Sharp upper bounds for sub-Gaussian maxima]
	Let $\{X_i\}_{i = 1}^n$ be a sequence of zero-mean sub-Gaussian variables with parameter $\sigma$. Then, we have
	\begin{equation}
		\bbE \off{\max_{i \in [n]} X_i} \leq \sqrt{2\sigma^2 \log n}, \quad \text{for all } n \geq 1.
	\end{equation}
	Note that independence assumptions are unnecessary.
\end{lem}

@@

\begin{rmk}
	The maxima for $n$ i.i.d.\ sub-Gaussian variables also share the similar concentration property as Gaussian variables in Exercise~\ref{Gausmax}. The maxima $\max_{i \in [n]} X_i$ concentrates around the sub-Gaussian parameter $\sigma$.
\end{rmk}

@@

@@

	\begin{rmk}
		The concentration around mean implies the concentration around median for $n$ variables, vice versa. 
	\end{rmk}

@@

	\begin{rmk}
		Kernel density estimator satisfies the bounded difference property and possesses a concentration property around the mean.
	\end{rmk}

@@

    \begin{rmk}
    	The concentration inequalities also apply to the Hilbert space. 
    \end{rmk}

@@

    \begin{rmk}
    	For a random vector $X$, Hanson-Wright Inequality implies the concentration property for the matrix product $X^T \mQ X$, where $\mQ$ is a positive semi-definite matrix.
    \end{rmk}

@@

    \begin{rmk}
    	For a random variable $X$, the existence of $X$'s Orlicz norm implies the boundedness of the tail probability of $X$.
    \end{rmk}

@@

\begin{rmk}
	The maxima of Orlicz variables concentrates around the Orlicz norm $\sigma$ and scales as $\psi^{-1}(n)$. In previous exercises, the maxima of sub-Gaussian variables concentrates around the sub-Gaussian variable $\sigma$ and scales at $\log (n)$. 
\end{rmk}

@@

 \begin{rmk}
 	For $n$ independent variables $X_i$, the moment condition for $X_i$ implies the boundedness of the tail probability for the sample mean $\frac{1}{n} \sum_{i=1}^n X_i$.
 \end{rmk}

@@

 \subsection{Exercise 2.22}

>>>>

	To evaluate $I_1$, we solve $1 = C \exp \of{-\frac{t^2}{4 \nu^2}} $, and the minimization term becomes
	\begin{align}
		\min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} = \begin{cases}
			\quad 1 & \text{when } t < 2\nu \sqrt{\log C},\\
			\quad C \exp \of{-\frac{t^2}{4 \nu^2}} & \text{when } t \geq 2\nu \sqrt{\log C}.
		\end{cases}
	\end{align}
	Therefore, let $ y = \frac{t}{2\nu} - \sqrt{\log C}$, we have 
	\begin{align}
		I_1 &= \int_{0}^{ 2\nu \sqrt{\log C}} 1 dt + \int_{2\nu \sqrt{\log C}}^{+\infty} C \exp \of{-\frac{t^2}{4 \nu^2}} dt\\
		&=  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2 - 2y \sqrt{\log C}} dy\\
		&\leq  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2} dy.
	\end{align}
	By Gaussian integral $\int_{0}^{+\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$, we obtain $I_1 \leq 2\nu (\sqrt{\pi} + \sqrt{\log C})$. Similarly, we evaluate $I_2$ as follows 
	\begin{align}
		I_2 &= \int_{0}^{4B \log C} 1 dt + \int_{4B \log C}^{+\infty} C \exp \of{-\frac{t}{4B}} dt\\
		 &= 4B (\log C + 1).
	\end{align}
	Hence, we obtain the expectation of $Z$,
	\begin{equation}
		\bbE[Z] = I_1 + I_2 \leq  2\nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (\log C + 1).
	\end{equation}

@@

The lower bound of $\bbE[Z_n] $ is
\begin{equation}\label{eq:e2112}
	\bbE[Z_n] \geq (1- 1/e) \sqrt{\sigma^2 \log n} , \quad \text{for all }n \geq 6.
\end{equation}

Moreover, we have
\begin{equation}\label{eq:e2113}
	\frac{\bbE[Z_n]}{\sqrt{2 \sigma^2 \log n}} \rightarrow 1, \quad \text{as $n \rightarrow +\infty$}.
\end{equation}

@@

\begin{proof}
	First, we prove the equation~\eqref{eq:e2111}.

	\vspace{0.2cm}
	Consider the tail probability of $Z_n$. We have
	\begin{equation}
		\bbP[Z_n \geq t] = \bbP[\max_{i \in [n]} |X_i| \geq t] = 1 - \bbP[\max_{i \in [n]} |X_i| < t] = 1 - (1- \bbP[|X_1| \geq t])^n,
	\end{equation}
	where the last equality follows from the independence of $\{X_i\}_{i=1}^n$. By Bernoulli's inequality, we have
	\begin{equation}
		(1- \bbP[|X_1| \geq t])^n \geq 1 - n \bbP[|X_1| \geq t], \quad \text{for all } t >0.
	\end{equation} 
	Therefore, the expectation of $Z_n$ follows
	\begin{equation}\label{eq:e211pro1}
		\bbE[Z_n] = \int_{0}^{+\infty} \bbP[Z_n \geq t] dt \leq  c +  \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt, \quad \text{for all } c > 0.
	\end{equation}
	 Since the tail bound $\bbP[U \geq t] \leq \sqrt{\frac{2}{\pi}} \frac{1}{t} e^{-t^2 /2}$ holds for all standard normal random variable $U$, the tail bound of $|X_1|$ satisfies
	 \begin{equation}\label{eq:e211proupper}
	 	\bbP[|X_1| \geq t] \leq 2 \sqrt{\frac{2}{\pi}} \frac{\sigma}{t} e^{-\frac{t^2}{2\sigma^2}}.
	 \end{equation}
	Let $ u = \frac{ t}{\sigma}$. Hence, for $c > \sigma$, the last integral in equation~\eqref{eq:e211pro1} is upper bounded as
	 \begin{align}
	 	 \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt &\leq \frac{2n \sigma}{c} \sqrt{\frac{2}{\pi}} \int_{c}^{+\infty} e^{-\frac{t^2}{2\sigma^2} }dt \\  
	 	 %& = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} e^{-\frac{u^2}{2}} dt \\
	 	 & \leq \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} u e^{-\frac{u^2}{2}} dt \\
	 	 & = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}  e^{-\frac{c^2}{2\sigma^2}}. \label{eq:e211pro2}
	 \end{align}
	 For all $n \geq 2$, let $c = \sqrt{2\sigma^2 \log n}$ and plug the $c$ into equations~\eqref{eq:e211pro1} and \eqref{eq:e211pro2}. Then, we obtain the result
	 \begin{equation}
	 	\bbE[Z_n] \leq \sqrt{2\sigma^2 \log n} + \frac{2\sigma}{\sqrt{2\log n}}   \sqrt{\frac{2}{\pi}}  \leq \sqrt{2\sigma^2 \log n} + \frac{4\sigma}{\sqrt{2\log n}}, \quad \text{for all }n \geq 2.
	 \end{equation}

	 \textbf{Alternative Way:} An alternative proof for equation~\eqref{eq:e2111} is below.

	 Since $\{X_i\}_{i=1}^n$ are i.i.d.\ random normal variables following $N(0, \sigma^2)$, by Exercise~\ref{ex:12}, we have 
	 \begin{equation}
	 	\bbE[\max_{i \in [n]} X_i] \leq \sqrt{2 \sigma^2 \log n}, \quad \text{for all } n \geq 1.
	 \end{equation}
	 Note that $\max_{i \in [n]} |X_i| = \max_{i \in [2n]} X_i$, where $X_{n + i} = - X_i $, for all $i\in[n]$. Then, we have $\max_{i \in [n]}|X_i| \leq \sqrt{2 \sigma^2 \log(2n)}$. Note that
	 \begin{equation}
	 	 \of{\sqrt{2 \sigma^2 \log n} + \frac{4\sigma}{\sqrt{2 \log n}}}^2 = 2 \sigma^2 \log n + 8\sigma^2 \of{1 + \frac{1}{\log n}} \geq 2 \sigma^2 \log n + 2 \sigma^2 \log 2 = (\sqrt{2 \sigma^2 \log 2n} )^2.
 	 \end{equation}
 	 Therefore, we obtain the upper bound
 	 \begin{equation}
 	 	\bbE[Z_n] \leq \sqrt{2 \sigma^2 \log(2n)} \leq \sqrt{2 \sigma^2 \log n} + \frac{4\sigma}{\sqrt{2 \log n}}.
 	 \end{equation}

@@

	 \vspace{0.2cm}
	 Last, we prove the convergence property~\eqref{eq:e2113}.

	\textbf{My thoughts:} To prove $\lim_{n \rightarrow +\infty}\frac{ \bbE[Z_n]}{\sqrt{2\sigma^2 \log n}}  = 1$, we need to prove $\limsup_{n \rightarrow +\infty}\frac{ \bbE[Z_n]}{\sqrt{2\sigma^2 \log n}}  = \liminf_{n \rightarrow +\infty}\frac{ \bbE[Z_n]}{\sqrt{2\sigma^2 \log n}} = 1$. The $\limsup$ part is easy because of the upper bound~\eqref{eq:e2111}. Then, we only need to prove the $\liminf$ part.

	Note that 
	\[\bbE[Z_n] \geq \sqrt{2 \sigma^2 \log n}  \bbP \off{Z_n > \sigma \sqrt{2 \log n}} =  \sigma \sqrt{2\log n} \off{1 - \of{1 - \bbP \off{|X_i| > \sigma \sqrt{2\log n}}   }^n}. \]

	If I can prove that $\lim_{n \rightarrow +\infty}\of{1 - \bbP \off{|X_i| > \sigma \sqrt{2\log n}}   }^n = 0 $, the problem is done.

	Note that $\bbP \off{|X_i| > \sigma \sqrt{2\log n}}  = 2 \bbP \off{X_i > \sigma \sqrt{2\log n}} \leq 2 \sqrt{\frac{2}{\pi}} \frac{\sigma}{\sqrt{2 \sigma^2 \log n}} e^{-\log n} = \frac{2}{\pi n \sqrt{\log n}}$. Then, the limit is upper bounded
	 \[\lim_{n \rightarrow +\infty}\of{1 - \bbP \off{|X_i| > \sigma \sqrt{2\log n}}   }^n \leq \lim_{n \rightarrow +\infty} \of{1 -  \frac{2}{\pi n \sqrt{\log n}} }^n = 1. \]

	 I can not get the result I want.

@@

\begin{rmk}
	The absolute Gaussian maxima $Z_n$ concentrates around $\sigma \sqrt{\log n}$ for large $n$. Standard Gaussian variables concentrate around 0, while the maxima of the $n$ standard Gaussian variables is no longer a Gaussian variable. The scale of Gaussian maxima scales as $\sqrt{n}$, however, the sum of $n$ i.i.d.\ Gaussian variables scales as $n$.
\end{rmk}

\subsection{Exercise 2.12}\label{ex:12}
\begin{lem}[Sharp upper bounds for sub-Gaussian maxima]
	Let $\{X_i\}_{i = 1}^n$ be a sequence of zero-mean sub-Gaussian variables with parameter $\sigma$. Then, we have
	\begin{equation}
		\bbE \off{\max_{i \in [n]} X_i} \leq \sqrt{2\sigma^2 \log n}, \quad \text{for all } n \geq 1.
	\end{equation}
	Note that independence assumptions are unnecessary.
\end{lem}

@@

\begin{rmk}
	The maxima for $n$ i.i.d.\ sub-Gaussian variables is upper bounded by $\sigma \sqrt{\log n}$. The independence assumption is not necessary for the upper bound of sub-Gaussian maxima.
\end{rmk}

@@

\vspace{0.2cm}
\textbf{My thoughts:} Note that $X_1 X_2 = \frac{1}{4}\of{(X_1 + X_2)^2 - (X_1 - X_2)^2}$. By part $(a)$, we have $X_1 + X_2$ and $X_1 - X_2$ are sub-Gaussian with the parameter $\sqrt{\sigma_1^2 + \sigma^2_2}$. Note that for a variable $X \sim \subG(\sigma^2)$, the variable $Z = X^2 - \bbE[X^2] \sim \text{subE}(16\sigma^2, 16\sigma^2)$. 
\vspace{0.2cm}

The characteristic function of $X_1 X_2$ is 
\begin{equation}\label{eq:e13c1}
	\bbE[e^{t X_1 X_2}] = \bbE[e^{\frac{t}{4} \of{(X_1 + X_2)^2 - (X_1 - X_2)^2} }] = \bbE[ e^{\frac{t}{4} (X_1 +X_2 )^2} e^{-\frac{t}{4} (X_1 - X_2)^2 } ], \quad \text{for all } t \in \bbR.
\end{equation} 
By Cauchy-Schwartz inequality, the expectation~\eqref{eq:e13c1} satisfies
\begin{equation}\label{eq:e13c2}
	\bbE[ e^{\frac{t}{4} (X_1 +X_2 )^2} e^{-\frac{t}{4} (X_1 - X_2)^2 } ] \leq \of{\bbE[ e^{\frac{t}{2} (X_1 +X_2 )^2} ]}^{1/2} \of{\bbE[ e^{-\frac{t}{2} (X_1 -X_2 )^2} ]}^{1/2}.
\end{equation}

Let $Z_1 = (X_1 + X_2)^2 - \bbE[(X_1 + X_2)^2]$. Note that $\bbE[X^2_i] = \var{(X_i)} \leq \sigma^2_i $, for $ i = 1,2$, and the variable $Z_1 \sim \text{subE}(16(\sigma^2_1 + \sigma^2_2), 16(\sigma^2_1 + \sigma^2_2))$. We have
\begin{equation}\label{eq:e13c3}
	\bbE[ e^{\frac{t}{2} (X_1 +X_2 )^2} ] = \bbE[e^{\frac{t}{2} Z_1 + \frac{t}{2} \bbE[(X_1 + X_2)^2] }] \leq \bbE[e^{\frac{t}{2} Z_1}] e^{ \frac{t}{2} (\sigma_1^2 + \sigma_2^2) } \leq e^{32 t^2 (\sigma^2_1 + \sigma_2^2)^2 + \frac{t}{2} (\sigma^2_1 + \sigma_2^2) },
\end{equation}
for all  $|t| \leq \frac{1}{8 (\sigma^2_1 + \sigma_2^2)}$. Similarly, we have
\begin{equation}\label{eq:e13c4}
	\bbE[ e^{-\frac{t}{2} (X_1 -X_2 )^2} ]  \leq  e^{32 t^2 (\sigma^2_1 + \sigma_2^2)^2 - \frac{t}{2} (\sigma^2_1 + \sigma_2^2) },
\end{equation}
for all  $|t| \leq \frac{1}{8 (\sigma^2_1 + \sigma_2^2)}$. Plug the inequalities \eqref{eq:e13c3} and \eqref{eq:e13c4} into the inequality~\eqref{eq:e13c2}. We have
\begin{equation}
		\bbE[ e^{\frac{t}{4} (X_1 +X_2 )^2} e^{-\frac{t}{4} (X_1 - X_2)^2 } ]  \leq  e^{32 t^2 (\sigma^2_1 + \sigma_2^2)^2}, \quad \text{for all } |t| \leq \frac{1}{8 (\sigma^2_1 + \sigma_2^2)}.
\end{equation}
Therefore, the variable $X_1 X_2 \sim \text{subE}(8(\sigma^2_1 + \sigma^2_2),8(\sigma^2_1 + \sigma^2_2) )$.

@@

	\begin{rmk}
	If a variable concentrates around mean, the variable also concentrates around the median, vice versa.
	\end{rmk}

@@

	\begin{rmk}
		Kernel density estimator satisfies the bounded difference property and possesses a concentration property. Kernel density estimator converges to its mean in probability at the speed of $e^{-n}$ when $n$ goes to infinity.
	\end{rmk}

@@

    \begin{rmk}
    	The sample mean also for $n$ variables in a Hilbert space converge to its expectation in probability at the speed of $e^{-n}$.
    \end{rmk}

@@

    \begin{rmk}
    	For a random vector $X$, Hanson-Wright Inequality implies the tail probability of matrix product $X^T \mQ X$ is upper bounded at the leve of $e^{-\min\{ \onorm{\mQ}^{-2}_F, \onorm{\mQ}^{-1}_{op} \}}$, where $\mQ$ is a positive semi-definite matrix. 
    \end{rmk}

@@

    \begin{rmk}
    	For a random variable $X$, the existence of $X$'s Orlicz norm implies the tail probability of $X$ is upper bounded at the level of $e^{-\onorm{X}^{-q}_{\varphi} }$. 
    \end{rmk}

@@

\begin{rmk}
	The maxima of Orlicz variables is upper bounded by the Orlicz norm $\sigma$ and scales as $\psi^{-1}(n)$. In previous exercises, the maxima of sub-Gaussian variables concentrates around the sub-Gaussian variable $\sigma$ and scales at $\log (n)$. 
\end{rmk}

@@

 \begin{rmk}
 	For $n$ independent zero-mean variables $X_i$, the order $2m$ moment condition for $X_i$ implies that the sample mean $\frac{1}{n} \sum_{i=1}^n X_i$ converges to 0 at the speed of $n^{-m}$.
 \end{rmk}

@@
